Optimal Finite Element Integration

### 2015-08-11 TOMS_Paper/assembly.tex
Several fixes to the seecond half of the paper

\subsection{Qualitative Comparison}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

\subsection{A-Priori Operation Counting}
\label{sec:op_count}
It remains to tie one loose hand: the construction of the pre-evaluation cost function. We introduce the cost function $\theta : M \to \mathbb{N} \times \mathbb{N}$ that, given a monomial, returns two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation. Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$. Assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.

For $\theta_{w}$, things are more complicated. We first need to account for the presence of (derivatives of) coefficients to estimate the \textit{increase factor} $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}.

@@

One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$. In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities. By expanding the product we observe an increase in the number of $L_k$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient). Intuitively, $\iota$ captures such a growth in non-hoistable operation.

With just a single coefficient, $\iota$ directly descends from the cost of expansion. In general, however, the calculation is less straightforward. Consider, for example, the case in which multiple coefficients originate from the same function space. Expansion would then lead to identical terms (i.e., identical pre-evaluated tables). Therefore, for a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are considering sets, so duplicates would be counted as 1).

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$. In particular, we need to calculate the number terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators employed. For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$. There are several ways of determining $\rho$. The fastest would be to extract it from high-level analysis of the operators in the form; for convenience, in our implementation we instead project (by analysis of the expression tree) the output of expansion and factorization in the monomial.

@@

\subsection{Automation through the COFFEE Language}
As opposed to what happens in FFC with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general: all optimizations are expressed as composition of parametric ``building-block'' operations. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could ideally be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler. Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.

A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion. These are parametric operations. For example, one could ask to factorize only constant or only linear terms, while hoisting can be limited by imposing a constraint on the size of the temporaries. Each of these operations is implemented by suitable manipulation of the abstract syntax tree representing the integration routine.

\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
\label{sec:heuristic-int-dep}
As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions. These are expressions that should logically be placed within $L_i$. They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point. For some complex monomials (e.g., see~\citeN{quadrature1} and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, outweigh that in the multilinear loop nest. In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.

Despite the fact that we are not characterizing optimality for this wider class of loops, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis}, particularly for removing sharing, to minimize the operational cost. This is straightforward in our code generation system by composing rewrite operators. Our strategy is as follows.

COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions. This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to trivial code without performing any sort of transformation. Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$. We then synthesize an optimal loop nest as described in the previous sections. This may in turn expose a set of $L_i$-dependent statements evaluating some temporaries. For each of this statements, we try to remove sharing by greedily applying factorization and code motion. In the COFFEE language, this process is expressed by simply concatenating five operators (i.e., five function calls named ``factorize'', ``expand'', etc.).

@@

\subsubsection{Review of Existing Optimizations}
\label{sec:opt-review}
We start with briefly reviewing the low-level optimizations presented in~\citeN{Luporini}.

@@

\paragraph{Vector-register Tiling} Blocking (or tiling) at the level of vector registers improves data locality beyond traditional unroll-and-jam transformations. The blocking strategy consists of evaluating outer products by just using two vector register without ever spilling to cache.

\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a relatively low number of logical registers (e.g. 16/32). By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in a separate loop nest.

\subsubsection{Vector-promotion of Integration-dependent Expressions}
\label{sec:vect-prom}
Integration-dependent expressions are inherently executed as scalar code because vectorization (unless employing special hand-written schemes) occurs along a single loop, typically the innermost. In our loop nest, $L_i$ is clearly an outer loop. In situations such as in Section~\ref{sec:heuristic-int-dep}, we may want to vectorize also along $L_i$. One way to achieve this is vector-promotion. This implies creating a clone loop of $L_i$ in the preheader of the main loop nest, in which vector temporaries, instead of scalar ones, are evaluated.

\subsubsection{Handling Block-sparse Tables}
\label{sec:zeros}
Consider a tabulated set of basis functions having quadrature points along the rows and functions along the columns. For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}. Unless using a smart form compiler (and one of our goals is precisely to relieve it from the burden of smart code generation), there are several circumstances in which the tables can be block-sparse. Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements. Zero-valued rows can result from using non-standard functions spaces, like Raviart-Thomas. Zero-valued blocks can appear in pre-evaluated temporaries. Our objective is a general system that avoids useless iteration over these zero-entries while preserving the effectiveness of the other low-level optimizations, especially vectorization.

In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g. \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed. Our approach, which will be compared to this pioneering work, aims to free the generated code from any indirection arrays. This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.

The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice (rows, columns) of zero-valued entries, and if that slice is smaller than the vector length, that we do nothing (i.e., the loop nest is not transformed). Otherwise, we restructure the loop nest. This has several complicated implications, including loop fission and a memory offsetting usually dangerous for padding and data alignment, which are, however, beyond the scope of this paper. The implementation is based on symbolic execution: the loop nests are traversed and for each statement the location of zeros in each of the involved symbols is tracked. Arithmetic operators have a different impact on the tracking. For example, for each loop dimension, multiplication requires computing the set intersection of the zero-valued slices, whereas addition requires computing the set union.

@@

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler's built-in optimization system: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler's built-in optimization system: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler's built-in optimization system: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, discussed in~\citeN{Fenics}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
\end{itemize}

@@

For a fair comparison, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants. UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake. Our patches fix this problem by producing code with a data storage layout as expected by Firedrake.

@@

Following the methodology adopted in ~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis are instead
\begin{itemize}
\item the space of test, trial, and coefficient functions, Lagrange
\item the mesh, tetrahedral with a total of 4374 elements
\item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
\end{itemize}

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are ``nested'' plots. We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''. In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over unmodified code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low-level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries. In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

>>>>

\subsection{Qualitative Comparison}
\label{sec:qualitative}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

\subsection{A-Priori Operation Counting}
\label{sec:op_count}
It remains to tie one loose hand: the construction of the pre-evaluation cost function. We introduce the cost function $\theta$ such that $\theta : M \to \mathbb{N} \times \mathbb{N}$; that is, given a monomial, two natural numbers representing the optimal operation count without ($\theta_{wo}$) and with ($\theta_{w}$) pre-evaluation are returned. Since $\theta$ is expected to be used by a compiler to drive the transformation process, requirements are simplicity and velocity.

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$. Assuming $I$ to be the size of the $L_i$ iteration space, we have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.

For $\theta_{w}$, things are more complicated. We first need to estimate the \textit{increase factor}, $\iota$, to account for the presence of (derivatives of) coefficients. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}.

@@

One can think of this as the (simplified) loop nest originating from the assembly of a pre-multiplied mass matrix. The sub-expression \texttt{f[0]*B[i,0]+f[1]*B[i,1]+f[2]*B[i,2]} represents the field $f$ over (tabulated) basis functions $B$. In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from the integration-dependent (i.e., $L_i$-dependent) quantities. By expanding the product we observe an increase in the number of $[L_j, L_k]$-dependent operations of a factor 3 (the local degrees of freedom for the coefficient). Intuitively, $\iota$ captures this growth in non-hoistable operation.

With a single coefficient, as we just saw, $\iota$ directly descends from the cost of expansion. In general, however, predicting $\iota$ is less straightforward. For example, consider the case in which a monomial has multiple coefficients expressed over the same function space. The expansion would now lead to identical sub-expressions that, once pre-evaluated, would be mapped to the same temporary. In other words, the resulting loop nest would be characterized by sharing. Therefore, the actual operation count (i.e., once sharing is removed) would be smaller than that one could infer from analysing the expansion ``in isolation''. For a precise estimate of $\iota$, we instead need to calculate the $k$-combinations with repetitions of $n$ elements, with $k$ being the number of coefficient-dependent terms appearing in a product (in the example, there is only $f$, so $k=1$)  and $n$ the cardinality of the set of symbols involved in the coefficient expansion (in the example, \texttt{B[i,0]}, \texttt{B[i,1]}, and \texttt{B[i,2]}, so $n=3$; note that we are talking about sets here, so duplicates would be counted once).

@@

If, on the other hand, $\iota < I$, a further step is necessary to estimate $\theta_{w}$. In particular, we need to calculate the number terms $\rho$ such that $\theta_{w} = \rho \cdot \iota$. Consider again Figure~\ref{code:increase_factor}. In the case of the mass matrix, the body of $L_k$ is simply characterized by the dot product of test and trial functions, \texttt{B[j]*B[k]}, so trivially $\rho = 1$. In general, $\rho$ varies with the discretization and the differential operators used in the form. For example, in the case of the bi-dimensional Poisson equation, after a suitable factorization, we have $\rho = 3$. There are several ways of determining $\rho$. The fastest would be to extract it from high-level analysis of form and discretization; for convenience, in our implementation we have algorithms that, based on analysis of the expression tree, project the output of monomial expansion and factorization, which in turn gives us $\rho$.

@@

\subsection{Automation through the COFFEE Language}
As opposed to what happens in the FEniCS Form Compiler with quadrature and tensor modes, there are no separate trunks in COFFEE handling pre-evaluation, sharing, and code motion in general. All optimizations are instead expressed as composition of parametric ``building-block'' operations. This has several advantages. Firstly, extendibility: novel transformations -- for instance, sum-factorization in spectral methods -- could be expressed using the existing operators, or with small effort building on what is already available. Secondly, generality: other domains sharing properties similar to that of finite element integration (e.g., multilinear loop nests) could be optimized through the same compiler. Thirdly, robustness: the same building-block operations are exploited, and therefore stressed, by different optimization pipelines.

A non-exhaustive list of such operations includes expansion, factorization, re-association, generalized code motion. These ``rewrite operators'' can be seen as the COFFEE language. They define parametric transformations: for example, one could ask to factorize constant rather linear terms, while hoisting could be driven by loop dependency. Their implementation is based on manipulation of the abstract syntax tree representing the integration routine.

\subsubsection{Heuristic Optimization of Integration-dependent Expressions}
\label{sec:heuristic-int-dep}
As a proof-of-concept of our generality claim, we briefly discuss our optimization strategy for integration-dependent expressions. These are expressions that should logically be placed within $L_i$. They can originate, for example, from the extensive use of tensor algebra in the derivation of the weak variational form or from the use of a non-affine reference-to-physical element mapping, which Jacobian needs be re-evaluated at every quadrature point. For some complex monomials and for coarser discretizations, the operation count within $L_i$ could be comparable or, in some circumstances, even outweigh that of the multilinear loop nest. In these cases, our definition of optimality becomes weaker, since the underlying assumption is that the bulk of the computation is carried out in innermost loops.

Despite the fact that we are not characterizing optimality for this wider class of problems, we can still heuristically apply the same reasoning of Sections~\ref{sec:optimal-impl} and~\ref{sec:optimal-synthesis} to remove sharing, thus reducing the operation count. This is straightforward in our code generation system by composing rewrite operators. Our strategy is as follows.

COFFEE is agnostic with respect to the high level form compiler, so the first step consists of removing redundant sub-expressions. This is because a form compiler abstracting from optimization will translate expressions as in Equation~\ref{eq:quadrature} directly to code without performing any sort of analysis. Eliminating redundant sub-expressions is usually helpful to relieve the arithmetic pressure inherent to $L_i$. We then synthesize an optimal loop nest as described in the previous sections. This may in turn expose a set of $L_i$-dependent expressions. For each of this expressions, we try to remove sharing by greedily applying factorization and code motion. In the COFFEE language, this process is expressed by simply composing five rewrite operators.

@@

\subsubsection{Review of Existing Optimizations}
\label{sec:opt-review}
We start with briefly reviewing the low level optimizations presented in~\citeN{Luporini}.

@@

\paragraph{Vector-register Tiling} Blocking (or tiling) at the level of vector registers improves data locality beyond traditional unroll-and-jam transformations. This blocking strategy consists of evaluating outer products by using just two vector register and without ever spilling to cache.

\paragraph{Expression Splitting} When the number of basis functions arrays (or, equivalently, temporaries introduced by code motion) and constants is large, the chances of spilling to cache are high in architectures with a few logical registers (e.g. 16/32). By exploiting sum’s associativity, an expression can be fissioned so that the resulting sub-expressions can be computed in separate loop nests. This reduces the register pressure.

\subsubsection{Vector-promotion of Integration-dependent Expressions}
\label{sec:vect-prom}
Integration-dependent expressions are inherently executed as scalar code because vectorization (unless employing special hand-written schemes) occurs along a single loop, typically the innermost. For the same reasons discussed in Section~\ref{sec:heuristic-int-dep}, we also want to vectorize along $L_i$. One way to achieve this is vector-promotion. This requires creating a ``clone'' of $L_i$ in the preheader of the loop nest, in which vector temporaries are evaluated in what is now an innermost loop.

\subsubsection{Handling Sparse Tables}
\label{sec:zeros}
Consider a set of tabulated basis functions with quadrature points along rows and functions along columns. For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}. Unless using a smart form compiler (which we want to avoid), there are circumstances in which the tables are sparse. Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements. Zero-valued rows can result from using non-standard functions spaces, like Raviart-Thomas. Zero-valued blocks can appear in pre-evaluated temporaries. Our objective is a transformation that avoids useless iteration over zeros while preserving the effectiveness of the other low-level optimizations, especially vectorization.

In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g. \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed. Our approach, which will be compared to this pioneering work, aims to free the generated code from such indirection arrays. This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.

The idea is that if the dimension along which vectorization is performed (typically the innermost) has a contiguous slice of zeros, but that slice is smaller than the vector length, then we do nothing (i.e., the loop nest is not transformed). Otherwise, we restructure the iteration space. This has several non-trivial implications. The most notable one is memory offsetting (e.g., \texttt{A[i+m,j+n]}), which dramatically enters in conflict with padding and data alignment. We use heuristics to retain the positive effects of both and to ensure correctness. Details are, however, beyond the scope of this paper.

The implementation is based on symbolic execution: the loop nests are traversed and for each statement encountered the location of zeros in each of the involved symbols is tracked. Arithmetic operators have a different impact on tracking. For example, multiplication requires computing the set intersection of the zero-valued slices (for each loop dimension), whereas addition requires computing the set union.

@@

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
\end{itemize}

@@

For the fairest comparison possible, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants. UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake. Our patches fix this problem by producing code with a data storage layout as expected by Firedrake.

@@

Following the methodology adopted in~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test and trial functions always have same degree
\item the polynomial order of coefficient (or ``pre-multiplying'') functions, $p \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis, instead, are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
\end{itemize}

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) can be interpreted as ``nested'' plots. We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''. In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.
